---
title: "DDA"
author: "Alireza Ghaharpour"
date: "2025-05-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 0. Instructions 

```{r}
# Check and install missing packages one by one

if (!require("dplyr")) install.packages("dplyr", dependencies = TRUE)
if (!require("ggplot2")) install.packages("ggplot2", dependencies = TRUE)
if (!require("validate")) install.packages("validate", dependencies = TRUE)
if (!require("tidyverse")) install.packages("tidyverse", dependencies = TRUE)
if (!require("Hmisc")) install.packages("Hmisc", dependencies = TRUE)
if (!require("psych")) install.packages("psych", dependencies = TRUE)
if (!require("e1071")) install.packages("e1071", dependencies = TRUE)
if (!require("VIM")) install.packages("VIM", dependencies = TRUE)
if (!require("corrplot")) install.packages("corrplot", dependencies = TRUE)
if (!require("RColorBrewer")) install.packages("RColorBrewer", dependencies = TRUE)
if (!require("caret")) install.packages("caret", dependencies = TRUE)
if (!require("plotly")) install.packages("plotly", dependencies = TRUE)
if (!require("cluster")) install.packages("cluster", dependencies = TRUE)
if (!require("stats")) install.packages("stats", dependencies = TRUE)
if (!require("tree")) install.packages("tree", dependencies = TRUE)
if (!require("randomForest")) install.packages("randomForest", dependencies = TRUE)
if (!require("kernlab")) install.packages("kernlab", dependencies = TRUE)
if (!require("class")) install.packages("class", dependencies = TRUE)
if (!require("smotefamily")) install.packages("smotefamily", dependencies = TRUE)
if (!require("MLmetrics")) install.packages("MLmetrics", dependencies = TRUE)
if (!require("ROCR")) install.packages("ROCR", dependencies = TRUE)
if (!require("xgboost")) install.packages("xgboost", dependencies = TRUE)
if (!require("Metrics")) install.packages("Metrics", dependencies = TRUE)

# Special case: install UBL from GitHub if not already installed
if (!require("UBL")) {
  if (!require("devtools")) install.packages("devtools", dependencies = TRUE)
  devtools::install_github("paobranco/UBL")
}

```

```{r}
# Add code here to load all the required libraries with `library()`.  
# Do not include any `install.package()` for any required packages in this rmd file.
library(dplyr)
library(ggplot2)
library(validate)
library(tidyverse)
library(Hmisc)
library(psych)
library(e1071)
library(VIM)
library(corrplot)
library(RColorBrewer)
library(caret)
library(plotly)
library(cluster)
library(stats)
library(tree)
library(randomForest)
library(kernlab)
library(class)
library(smotefamily)
library(MLmetrics)
library(ROCR)
library(xgboost)
library(UBL)
library(Metrics)


```
## Function
```{r}
MinMax <- function(x) {
  if (is.numeric(x) && length(unique(x)) > 1) {
    return((x - min(x)) / (max(x) - min(x)))
  } else {
    return(x)
  }
}
```


# 1. Organise and clean the data

## 1.1 Data quality plan
```{r}
# Load the dataset
financial_df <- read.csv('financial_risk_assessment.csv')

# Get the structure of the dataset (data types and first few rows of each column)
str(financial_df) 

# Get summary statistics for all variables in the dataset
summary(financial_df) 

# Gender Encoding
financial_df$Gender <- as.factor(financial_df$Gender)

# Education Level Encoding
financial_df$Education.Level <- as.factor(financial_df$Education.Level)

# Marital Status Encoding
financial_df$Marital.Status <- as.factor(financial_df$Marital.Status)

# Loan Purpose Encoding
financial_df$Loan.Purpose <- as.factor(financial_df$Loan.Purpose)

# Employment Status Encoding
financial_df$Employment.Status <- as.factor(financial_df$Employment.Status)

# Payment History Encoding
financial_df$Payment.History <- as.factor(financial_df$Payment.History)

# Risk Rating Encoding
financial_df$Risk.Rating <- as.factor(financial_df$Risk.Rating)

```
Our dataset consists of 20 variables (columns) and 15,000 rows (attributes).

The dataset contains two types of variables:

. Numerical variables: Age, Income, Credit Score, Loan Amount, Years at Current Job, Debt-to-Income Ratio, Assets Value, Number of Dependents, Previous Defaults, Marital Status Change.

. Categorical variables: Gender, Education Level, Marital Status, Loan Purpose, Employment Status, Payment History, City, State, Country, Risk Rating.


## 1.2 Data quality findings
```{r}
# Check for missing values in each column
colSums(is.na(financial_df))

sum(duplicated(financial_df))

# Frequency tables for categorical variables

table(financial_df$Gender)
table(financial_df$Education.Level)
table(financial_df$Marital.Status)
table(financial_df$Loan.Purpose)
table(financial_df$Employment.Status)
table(financial_df$Payment.History)
#table(financial_df$City)
#table(financial_df$State)
#table(financial_df$Country)
table(financial_df$Risk.Rating)


```


There are no missing values in the categorical variables, and they appear to be correctly formatted. However, some numerical columns contain missing values (NA’s), affecting approximately 15% of the dataset. These missing values need to be cleaned before further analysis.

## 1.3 Data cleaning


According to "COMPARISON OF SIMPLE MISSING DATA IMPUTATION TECHNIQUES FOR NUMERICAL AND CATEGORICAL DATASETS" Article, When less than 20% of data is missing, mean and median imputations are effective in regression problems. kNN imputation is better at 20% missingness and significantly better when 50% or more data is missing. so 2250 form 15000 observation are missing so it is less than 20% of data and we use mean or median imputation

### 1.3.1 Remove location based variables

```{r}
financial_df_cleaned <- financial_df[, !(names(financial_df) %in% c('City' , 'State', 'Country'))]
str(financial_df_cleaned)
```
### 1.3.2 Remove NA's

```{r}
##############Distribution of Income
# More complete histogram with density curve
hist(financial_df_cleaned$Income, 
     breaks=60, 
     probability=TRUE, 
     col="lightblue",
     main="Distribution of Income",
     xlab="Income")
lines(density(financial_df_cleaned$Income,na.rm = TRUE), col="red", lwd=2)

# Add mean line
abline(v=mean(financial_df_cleaned$Income, na.rm=TRUE), 
       col="darkblue", lwd=2, lty=2)

# Add median line
abline(v=median(financial_df_cleaned$Income, na.rm=TRUE), 
       col="darkgreen", lwd=2, lty=3)


# Check the skewness of Income in Numbers
skewness(financial_df_cleaned$Income, na.rm = TRUE)


##############Distribution of Credit.Score
# More complete histogram with density curve
hist(financial_df_cleaned$Credit.Score, 
     breaks=20, 
     probability=TRUE, 
     col="lightblue",
     main="Distribution of Credit.Score",
     xlab="Credit.Score")
lines(density(financial_df_cleaned$Credit.Score,na.rm = TRUE), col="red", lwd=2)

# Add mean line
abline(v=mean(financial_df_cleaned$Credit.Score, na.rm=TRUE), 
       col="darkblue", lwd=2, lty=2)

# Add median line
abline(v=median(financial_df_cleaned$Credit.Score, na.rm=TRUE), 
       col="darkgreen", lwd=2, lty=3)


# Check the skewness of Credit.Score in Numbers
skewness(financial_df_cleaned$Credit.Score, na.rm = TRUE)


##############Distribution of Loan.Amount
# More complete histogram with density curve
hist(financial_df_cleaned$Loan.Amount, 
     breaks=20, 
     probability=TRUE, 
     col="lightblue",
     main="Distribution of Loan.Amount",
     xlab="Loan.Amount")
lines(density(financial_df_cleaned$Loan.Amount,na.rm = TRUE), col="red", lwd=2)

# Add mean line
abline(v=mean(financial_df_cleaned$Loan.Amount, na.rm=TRUE), 
       col="darkblue", lwd=2, lty=2)

# Add median line
abline(v=median(financial_df_cleaned$Loan.Amount, na.rm=TRUE), 
       col="darkgreen", lwd=2, lty=3)

# Check the skewness of Loan.Amount in Numbers
skewness(financial_df_cleaned$Loan.Amount, na.rm = TRUE)


##############Distribution of Assets.Value
# More complete histogram with density curve
hist(financial_df_cleaned$Assets.Value, 
     breaks=20, 
     probability=TRUE, 
     col="lightblue",
     main="Distribution of Assets.Value",
     xlab="Assets.Value")
lines(density(financial_df_cleaned$Assets.Value,na.rm = TRUE), col="red", lwd=2)

# Add mean line
abline(v=mean(financial_df_cleaned$Assets.Value, na.rm=TRUE), 
       col="darkblue", lwd=2, lty=2)

# Add median line
abline(v=median(financial_df_cleaned$Assets.Value, na.rm=TRUE), 
       col="darkgreen", lwd=2, lty=3)

# Check the skewness of Assets.Value in Numbers
skewness(financial_df_cleaned$Assets.Value, na.rm = TRUE)


```


Since all distributions are not skewed, Mean Imputation is recommended. If the distributions were skewed, Median Imputation would have been a better option.


```{r}
# Impute missing values with the mean for continuous variables
financial_df_cleaned$Income[is.na(financial_df_cleaned$Income)] <- mean(financial_df_cleaned$Income, na.rm = TRUE)

financial_df_cleaned$Credit.Score[is.na(financial_df_cleaned$Credit.Score)] <- mean(financial_df_cleaned$Credit.Score, na.rm = TRUE)

financial_df_cleaned$Loan.Amount[is.na(financial_df_cleaned$Loan.Amount)] <- mean(financial_df_cleaned$Loan.Amount, na.rm = TRUE)

financial_df_cleaned$Assets.Value[is.na(financial_df_cleaned$Assets.Value)] <- mean(financial_df_cleaned$Assets.Value, na.rm = TRUE)

```

```{r}
financial_df_cleaned <- kNN(financial_df_cleaned, variable = c('Previous.Defaults', 'Number.of.Dependents'), k=5)

# Remove the _imp indicator columns
financial_df_cleaned <- financial_df_cleaned[, !grepl("_imp", names(financial_df_cleaned))]

# Display summary statistics after imputation
str(financial_df_cleaned) 

colSums(is.na(financial_df_cleaned))

```

# 2. EDA

## 2.1 Graphical Analysis
```{r}

# Selecting only numerical columns for Linear corrleation
numeric_df <- financial_df_cleaned[, sapply(financial_df_cleaned, is.numeric)]

# Creating a correlation matrix and scatterplot matrix
psych::pairs.panels(numeric_df, 
                    method = "pearson",  # Correlation method
                    hist.col = "lightblue",  # Histogram color
                    density = TRUE,  # Show density plot
                    ellipses = TRUE,  # Confidence ellipses
                    pch = 21, bg = "lightgray")  # Point style

```

```{r}
# Convert Risk Rating to numeric for Spearman correlation 
risk_numeric <- as.numeric(financial_df_cleaned$Risk.Rating)
# Compute Spearman correlation between Risk Rating and all numeric variables
Risk_correlations <- sapply(numeric_df, function(x){
  cor(risk_numeric, x, method = "spearman", use = "complete.obs")
})
# Create a data frame from the correlation results
corr_df <- data.frame(Variable = names(Risk_correlations), spearman_correlation = round(Risk_correlations, 3))
# Set scale limits for heatmap
min_val <- min(corr_df$spearman_correlation)
max_val <- max(corr_df$spearman_correlation)
#Plot Heatmap to show the correlations
ggplot(corr_df, aes(x = reorder(Variable, spearman_correlation),
                    y = "Risk Rating",fill = spearman_correlation)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(spearman_correlation, 2)),color = "black", size = 4, fontface = "bold") +
  scale_fill_gradient2(low = "yellow", mid = "blue", high = "orange",
                       midpoint = 0, limits = c(min_val, max_val),name = "Spearman\nCorrelation") +
  labs(title = "Correlation of Risk Rating with Numerical Features",x = "Variables", y = "") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),plot.title = element_text(face = "bold", hjust = 0.5))
```


```{r}
numeric_vars <- c("Age", "Income", "Credit.Score", "Loan.Amount", "Years.at.Current.Job", 
                  "Debt.to.Income.Ratio", "Assets.Value", "Number.of.Dependents", "Previous.Defaults", "Marital.Status.Change")


for (var in numeric_vars) {
  print(
    ggplot(financial_df_cleaned, aes(x=Risk.Rating, y=.data[[var]], fill=Risk.Rating)) +
      geom_boxplot() +
      theme_minimal() +
      labs(title=paste("Boxplot of", var, "by Risk Rating"), 
           x="Risk Rating", y=var)
  )
}



```

## 2.2 Unsupervised Learning Method

### 2.2.1 PCA 
```{r}
# Create dummy encoder for all predictors (excluding target)
dummies <- dummyVars(~ ., data = financial_df_cleaned[, names(financial_df_cleaned) != "Risk.Rating"])

# Apply transformation
financial_encoded <- predict(dummies, newdata = financial_df_cleaned)

# Convert to dataframe
financial_encoded_df <- as.data.frame(financial_encoded)

#Standardizing the data so that all variables contribute equally.
scaled_df <- scale(financial_encoded_df)

#Apply PCA
pca_result <- prcomp(scaled_df, center = TRUE, scale. = TRUE)

# View summary of PCA
summary(pca_result)

###  calculate the proportion of explained variance (PEV) from the std values
pca_variance <- pca_result$sdev^2
pca_var_percent <- round(pca_variance / sum(pca_variance) * 100, 1)

barplot(pca_var_percent,
        names.arg = paste0("PC", 1:length(pca_var_percent)),
        main = "Explained Variance by Principal Component",
        xlab = "Principal Components", ylab = "Variance Explained (%)",
        col = "lightblue")

###  plot the cumulative PEV
{
plot(cumsum(pca_var_percent), type = "b", pch = 19,
     xlab = "Number of Principal Components",
     ylab = "Cumulative Variance Explained (%)",
     main = "Cumulative Explained Variance")
abline(h = 80, col = "red", lty = 2)  # Optional: line at 80%
}

#Plot PC1 Loadings 
pca1_loadings <- pca_result$rotation[, 1]
variable <- rownames(pca_result$rotation)

# Create a dataframe
loadings_df <- data.frame(
  Variables = variable,
  Loading = pca1_loadings,
  Label = variable
)

p <- ggplot(loadings_df, aes(x = reorder(Label, Loading), y = Loading,
                             text = paste(Variables, "<br>Loading:", round(Loading, 3)))) +
  geom_col(fill = "red") +
  labs(title = "PC1 Loadings",x = "PC1",y = "Loadings Value") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"))

#add message on hover when user hover on the bar it sees message shows variable name and loading value by using ggplotly library 
interactive_plot <- ggplotly(p, tooltip = "text")
interactive_plot
```
### 2.2.2 Hierarchical Clustering

```{r}
###  hierarchical clustering - complete linkage
###   note: exclude the Risk Rating
dist_financial_clean <- dist(financial_encoded_df, method = 'euclidian')
hc_financial_clean <- hclust(dist_financial_clean, method = 'complete')
### 2.2 plot the associated dendrogram
plot(hc_financial_clean, hang = -0.1, labels = financial_df_cleaned$Risk.Rating)

### 2.3 select a partition containing 3 groups
hc_cluster_id_financial_clean <- cutree(hc_financial_clean, k = 3)
### 2.4 k-means with 3 groups
k_financial_clean = kmeans(financial_encoded_df, 3)
k_cluster_id_financial_clean <- k_financial_clean$cluster
```

#### 2.2.2.1 Evaluation of cluster results


```{r}
### 3.1 silhoutte score
sil_hc_financial_clean <- cluster::silhouette(hc_cluster_id_financial_clean, dist_financial_clean)
sil_k_financial_clean <- cluster::silhouette(k_cluster_id_financial_clean, dist_financial_clean)
### 3.2 silhoutte plots
###   note: use border = 'grey' to be able to see the plot lines
opar <- par(no.readonly = TRUE)
par(mfrow = c(2,1))
plot(sil_hc_financial_clean, border = 'grey')
plot(sil_k_financial_clean, border = 'grey')
```


K-means appears to be the better clustering method in this case for several reasons:

. Higher average silhouette width (0.45 vs 0.37) - indicating better overall cluster separation and cohesion

. More consistent silhouette widths across clusters (all around 0.45-0.46) - suggesting uniform quality of clustering

. The lowest silhouette width in K-means (0.45) is still higher than the average in HC (0.37)

In hierarchical clustering, there's more variability between clusters (0.29 to 0.51), suggesting uneven cluster quality.


#3 Supervised learning




## 3.1 Data Separation

```{r}
# set random seed
set.seed(1996)
# create a 70/30 training/test set split
n_rows <- nrow(financial_df_cleaned)
# sample 70% (n_rows * 0.7) indices in the ranges 1:nrows
training_idx <- sample(n_rows, n_rows * 0.7)
# filter the data frame with the training indices (and the complement)
training_financial <- financial_df_cleaned[training_idx,]
test_financial <- financial_df_cleaned[-training_idx,]
```



## 3.1 SVM

### 3.1.1 SVM Pure

```{r}

svm_formula <- reformulate(names(training_financial[, -17]), response = "Risk.Rating")

svm_model <- ksvm(svm_formula, data = training_financial, kernel = "rbfdot", C = 1)
svm_pred <- predict(svm_model, test_financial[, -17])

confusionMatrix(svm_pred, test_financial$Risk.Rating)

```

###3.1.2 SVM with Clustering
```{r}
# Step 1: Add cluster labels to the data
financial_df_clustered <- cbind(financial_df_cleaned, Cluster = as.factor(k_cluster_id_financial_clean))

# Step 2: Train-Test Split (as before, assuming train-test index is already prepared)
training_financial_clustered <- financial_df_clustered[training_idx, ]
test_financial_clustered <- financial_df_clustered[-training_idx, ]

# Step 3: Create formula for SVM
financial_data_formula_clustered <- reformulate(names(training_financial_clustered[, -17]), response = 'Risk.Rating')

# Step 4: Train SVM with linear and RBF kernels
rbf_svm_clustered <- ksvm(financial_data_formula_clustered, data = training_financial_clustered, kernel = 'rbfdot', C = 1)

# Step 5: Predict on the test data
rbf_svm_clustered_pred <- predict(rbf_svm_clustered, test_financial_clustered[,-17])

# Step 6: Confusion Matrix
confusionMatrix(rbf_svm_clustered_pred, test_financial_clustered$Risk.Rating)

```

###3.1.3 SVM One SMOTE and UnderSampling

```{r}
# Step 1: Combine features and label
df_knn3 <- cbind(Risk.Rating = financial_df_cleaned$Risk.Rating, financial_encoded_df)

# Step 2: Convert label to factor and integer for SMOTE
df_knn3$Risk.Rating <- as.factor(df_knn3$Risk.Rating)
df_knn3$Risk.Rating <- as.integer(df_knn3$Risk.Rating)  # SMOTE نیاز به عددی داره

# Step 3: Apply SMOTE
smote_result3 <- SMOTE(X = df_knn3[,-1], target = df_knn3$Risk.Rating, K = 5 , dup_size = 2)

# Step 4: Extract data
df_balanced_smote3 <- smote_result3$data
colnames(df_balanced_smote3)[ncol(df_balanced_smote3)] <- "Risk.Rating"
df_balanced_smote3$Risk.Rating <- factor(df_balanced_smote3$Risk.Rating,
                                        labels = levels(financial_df_cleaned$Risk.Rating))

table(df_balanced_smote3$Risk.Rating)

# undersample

df_balanced_undersampled <- df_balanced_smote3 %>%
  group_by(Risk.Rating) %>%
  slice_sample(n = 4500) %>%
  ungroup()

table(df_balanced_undersampled$Risk.Rating)

# Step 5: Normalize (optional)
df_balanced_undersampled[,-ncol(df_balanced_undersampled)] <- as.data.frame(lapply(df_balanced_undersampled[,-ncol(df_balanced_undersampled)], MinMax))

# Step 6: Train/Test split
train_idx <- sample(nrow(df_balanced_undersampled), 0.7 * nrow(df_balanced_undersampled))
train_data <- df_balanced_undersampled[train_idx, ]
test_data <- df_balanced_undersampled[-train_idx, ]

# Step 7: SVM

names(train_data) <- make.names(names(train_data))
names(test_data) <- make.names(names(test_data))
svm_formula <- reformulate(names(train_data[,-ncol(train_data)]), response = "Risk.Rating")

svm_model_rbf <- ksvm(svm_formula, data = train_data, kernel = "rbfdot", C = 1)
svm_pred <- predict(svm_model_rbf, test_data[,-ncol(test_data)])

# Step 8: Confusion Matrix

confusionMatrix(svm_pred, test_data$Risk.Rating)


svm_pred <- factor(svm_pred, levels = levels(test_data$Risk.Rating))
truth <- test_data$Risk.Rating

# List of classes
classes <- levels(truth)

# Compute precision, recall, and F1 score for each class
for (cls in classes) {
  cat(paste0("\nClass: ", cls, "\n"))
  precision <- Precision(y_pred = svm_pred, y_true = truth, positive = cls)
  recall <- Recall(y_pred = svm_pred, y_true = truth, positive = cls)
  f1 <- F1_Score(y_pred = svm_pred, y_true = truth, positive = cls)
  
  cat(paste("  Precision:", round(precision, 4), "\n"))
  cat(paste("  Recall:   ", round(recall, 4), "\n"))
  cat(paste("  F1 Score: ", round(f1, 4), "\n"))
}


```


Show ROC Curve


```{r}
train_data$Risk.Rating <- as.factor(train_data$Risk.Rating)
test_data$Risk.Rating  <- as.factor(test_data$Risk.Rating)

svm_model_rbf_prob <- ksvm(
  svm_formula,
  data = train_data,
  kernel = "rbfdot",
  C = 1,
  prob.model = TRUE
)

# Use model.frame to ensure feature compatibility with the model
test_features <- model.frame(svm_formula, data = test_data)

# Get probability matrix for each class
svm_prob <- predict(svm_model_rbf_prob, newdata = test_features, type = "probabilities")

# Initial check
head(svm_prob)

true_labels <- test_data$Risk.Rating
classes <- levels(true_labels)
colors <- c("red", "blue", "darkgreen")

# Start plotting
plot(NULL, xlim = c(0,1), ylim = c(0,1), xlab = "False Positive Rate", ylab = "True Positive Rate",
     main = "SVM ROC Curve - SMOTE - UnderSampling")

# Draw ROC curve for each class
for (i in seq_along(classes)) {
  cls <- classes[i]
  
  # Create prediction and performance objects for class i
  pred <- prediction(predictions = svm_prob[, cls], labels = as.numeric(true_labels == cls))
  perf <- performance(pred, "tpr", "fpr")
  
  plot(perf, add = TRUE, col = colors[i], lwd = 2)
}

abline(a = 0, b = 1, col = "gray", lty = 2)
legend("bottomright", legend = classes, col = colors, lty = 1, lwd = 2, bty = "n")


```


###3.1.4 SVM 2SMOTE

```{r}

# Step 1: Combine features and label
df_knn <- cbind(Risk.Rating = financial_df_cleaned$Risk.Rating, financial_encoded_df)

# Step 2: Convert label to factor and integer for SMOTE
df_knn$Risk.Rating <- as.factor(df_knn$Risk.Rating)
df_knn$Risk.Rating <- as.integer(df_knn$Risk.Rating)  # SMOTE requires numeric target

# Step 3: Apply SMOTE
smote_result <- SMOTE(X = df_knn[,-1], target = df_knn$Risk.Rating, K = 5 , dup_size = 5)

# Step 4: Extract data
df_balanced_smote <- smote_result$data
colnames(df_balanced_smote)[ncol(df_balanced_smote)] <- "Risk.Rating"
df_balanced_smote$Risk.Rating <- factor(df_balanced_smote$Risk.Rating,
                                        labels = levels(financial_df_cleaned$Risk.Rating))

# Convert label to integer again (required for SMOTE)
df_balanced_smote$Risk.Rating <- as.integer(df_balanced_smote$Risk.Rating)

# Apply second round of SMOTE on the previously SMOTE-balanced dataset
smote_result_2 <- SMOTE(X = df_balanced_smote[,-ncol(df_balanced_smote)],
                        target = df_balanced_smote$Risk.Rating,
                        K = 5,
                        dup_size = 1)

# Prepare final output
df_balanced_smote_2 <- smote_result_2$data
colnames(df_balanced_smote_2)[ncol(df_balanced_smote_2)] <- "Risk.Rating"

# Convert label back to factor with original levels
df_balanced_smote_2$Risk.Rating <- factor(df_balanced_smote_2$Risk.Rating,
                                          labels = levels(financial_df_cleaned$Risk.Rating))

# Check final class distribution
table(df_balanced_smote_2$Risk.Rating)

# Step 5: Normalize (optional)
df_balanced_smote_2[,-ncol(df_balanced_smote_2)] <- as.data.frame(lapply(df_balanced_smote_2[,-ncol(df_balanced_smote_2)], MinMax))

# Step 6: Train/Test split
train_idx <- sample(nrow(df_balanced_smote_2), 0.7 * nrow(df_balanced_smote_2))
train_data <- df_balanced_smote_2[train_idx, ]
test_data <- df_balanced_smote_2[-train_idx, ]

# Step 7: SVM
names(train_data) <- make.names(names(train_data))
names(test_data) <- make.names(names(test_data))
svm_formula <- reformulate(names(train_data[,-ncol(train_data)]), response = "Risk.Rating")

svm_model_rbf <- ksvm(svm_formula, data = train_data, kernel = "rbfdot", C = 1)
svm_pred <- predict(svm_model_rbf, test_data[,-ncol(test_data)])

# Step 8: Confusion Matrix
confusionMatrix(svm_pred, test_data$Risk.Rating)

# Convert to factor with consistent levels
svm_pred <- factor(svm_pred, levels = levels(test_data$Risk.Rating))
truth <- test_data$Risk.Rating

# List of classes
classes <- levels(truth)

# Compute precision, recall, and F1 score for each class
for (cls in classes) {
  cat(paste0("\nClass: ", cls, "\n"))
  precision <- Precision(y_pred = svm_pred, y_true = truth, positive = cls)
  recall <- Recall(y_pred = svm_pred, y_true = truth, positive = cls)
  f1 <- F1_Score(y_pred = svm_pred, y_true = truth, positive = cls)
  
  cat(paste("  Precision:", round(precision, 4), "\n"))
  cat(paste("  Recall:   ", round(recall, 4), "\n"))
  cat(paste("  F1 Score: ", round(f1, 4), "\n"))
}



```

Show ROC Curve


```{r}
train_data$Risk.Rating <- as.factor(train_data$Risk.Rating)
test_data$Risk.Rating  <- as.factor(test_data$Risk.Rating)

# Ensure that the target column is a factor with 3 levels
levels(train_data$Risk.Rating)


svm_model_rbf_prob <- ksvm(
  svm_formula,
  data = train_data,
  kernel = "rbfdot",
  C = 1,
  prob.model = TRUE
)

# Use model.frame to ensure compatibility of columns with the model
test_features <- model.frame(svm_formula, data = test_data)

# Get the probability matrix for each class
svm_prob <- predict(svm_model_rbf_prob, newdata = test_features, type = "probabilities")

# Initial check
head(svm_prob)

true_labels <- test_data$Risk.Rating
classes <- levels(true_labels)
colors <- c("red", "blue", "darkgreen")

# Start plot
plot(NULL, xlim = c(0,1), ylim = c(0,1), xlab = "False Positive Rate", ylab = "True Positive Rate",
     main = "SVM ROC Curve - 2SMOTE")

# Plot ROC for each class
for (i in seq_along(classes)) {
  cls <- classes[i]
  
  # Create prediction and performance objects for class i
  pred <- prediction(predictions = svm_prob[, cls], labels = as.numeric(true_labels == cls))
  perf <- performance(pred, "tpr", "fpr")
  
  plot(perf, add = TRUE, col = colors[i], lwd = 2)
}

abline(a = 0, b = 1, col = "gray", lty = 2)
legend("bottomright", legend = classes, col = colors, lty = 1, lwd = 2, bty = "n")


```



###3.1.5 SMOTE and merge High and Med (binary SVM)
```{r}
# Step 1: Combine features and label
df_knn4 <- cbind(Risk.Rating = financial_df_cleaned$Risk.Rating, financial_encoded_df)

# Step 2: Convert label to factor and integer for SMOTE
df_knn4$Risk.Rating <- as.factor(df_knn4$Risk.Rating)
df_knn4$Risk.Rating <- as.integer(df_knn4$Risk.Rating)  # SMOTE requires numeric labels

# Step 3: Apply SMOTE
smote_result4 <- SMOTE(X = df_knn4[,-1], target = df_knn4$Risk.Rating, K = 5 , dup_size = 2)

# Step 4: Extract data
df_balanced_smote4 <- smote_result4$data
colnames(df_balanced_smote4)[ncol(df_balanced_smote4)] <- "Risk.Rating"
df_balanced_smote4$Risk.Rating <- factor(df_balanced_smote4$Risk.Rating,
                                         labels = levels(financial_df_cleaned$Risk.Rating))

# Check class distribution
table(df_balanced_smote4$Risk.Rating)

# Convert to binary: "Low" vs "MedHigh"
df_binary <- df_balanced_smote4 %>%
  mutate(Risk.Rating = ifelse(Risk.Rating == "Low", "Low", "MedHigh")) %>%
  mutate(Risk.Rating = factor(Risk.Rating))

# Check new class distribution
table(df_binary$Risk.Rating)

# Normalize features
df_binary[, -ncol(df_binary)] <- as.data.frame(lapply(df_binary[, -ncol(df_binary)], MinMax))

# Train/Test split
set.seed(1996)
train_idx <- sample(nrow(df_binary), 0.7 * nrow(df_binary))
train_data <- df_binary[train_idx, ]
test_data <- df_binary[-train_idx, ]

# Clean variable names for formula
names(train_data) <- make.names(names(train_data))
names(test_data) <- make.names(names(test_data))

# Create formula
svm_formula <- reformulate(names(train_data[, -ncol(train_data)]), response = "Risk.Rating")

# Train binary SVM model
svm_model_binary <- ksvm(svm_formula, data = train_data, kernel = "rbfdot", C = 1)

# Predict and evaluate
svm_pred <- predict(svm_model_binary, test_data[, -ncol(test_data)])
confusionMatrix(svm_pred, test_data$Risk.Rating)

# Ensure factor levels match
svm_pred <- factor(svm_pred, levels = levels(test_data$Risk.Rating))
truth <- test_data$Risk.Rating

# List of classes
classes <- levels(truth)

# Compute precision, recall, and F1 score for each class
for (cls in classes) {
  cat(paste0("\nClass: ", cls, "\n"))
  precision <- Precision(y_pred = svm_pred, y_true = truth, positive = cls)
  recall <- Recall(y_pred = svm_pred, y_true = truth, positive = cls)
  f1 <- F1_Score(y_pred = svm_pred, y_true = truth, positive = cls)
  
  cat(paste("  Precision:", round(precision, 4), "\n"))
  cat(paste("  Recall:   ", round(recall, 4), "\n"))
  cat(paste("  F1 Score: ", round(f1, 4), "\n"))
}


```


Show ROC Curve


```{r}
train_data$Risk.Rating <- as.factor(train_data$Risk.Rating)
test_data$Risk.Rating  <- as.factor(test_data$Risk.Rating)

svm_model_rbf_prob <- ksvm(
  svm_formula,
  data = train_data,
  kernel = "rbfdot",
  C = 1,
  prob.model = TRUE
)

# Use model.frame to ensure column compatibility with the model
test_features <- model.frame(svm_formula, data = test_data)

# Get probability matrix for each class
svm_prob <- predict(svm_model_rbf_prob, newdata = test_features, type = "probabilities")

# Initial inspection
head(svm_prob)

true_labels <- test_data$Risk.Rating
classes <- levels(true_labels)
colors <- c("red", "blue")

# Start plot
plot(NULL, xlim = c(0,1), ylim = c(0,1), xlab = "False Positive Rate", ylab = "True Positive Rate",
     main = "SVM ROC Curve - Merge Class")

# Plot ROC curve for each class
for (i in seq_along(classes)) {
  cls <- classes[i]
  
  # Create prediction and performance object for class i
  pred <- prediction(predictions = svm_prob[, cls], labels = as.numeric(true_labels == cls))
  perf <- performance(pred, "tpr", "fpr")
  
  plot(perf, add = TRUE, col = colors[i], lwd = 2)
}

abline(a = 0, b = 1, col = "gray", lty = 2)
legend("bottomright", legend = classes, col = colors, lty = 1, lwd = 2, bty = "n")

```

#SVM +CrossValidation + Tuning

```{r}


# Step 1: Combine features and label
df_knn <- cbind(Risk.Rating = financial_df_cleaned$Risk.Rating, financial_encoded_df)

# Step 2: Convert label to factor and integer for SMOTE
df_knn$Risk.Rating <- as.factor(df_knn$Risk.Rating)
df_knn$Risk.Rating <- as.integer(df_knn$Risk.Rating)

# Step 3: Apply SMOTE
smote_result <- SMOTE(X = df_knn[,-1], target = df_knn$Risk.Rating, K = 5 , dup_size = 5)

# Step 4: Second round of SMOTE
df_balanced_smote <- smote_result$data
colnames(df_balanced_smote)[ncol(df_balanced_smote)] <- "Risk.Rating"
df_balanced_smote$Risk.Rating <- as.integer(df_balanced_smote$Risk.Rating)

smote_result_2 <- SMOTE(X = df_balanced_smote[,-ncol(df_balanced_smote)],
                        target = df_balanced_smote$Risk.Rating,
                        K = 5, dup_size = 1)

df_balanced_smote_2 <- smote_result_2$data
colnames(df_balanced_smote_2)[ncol(df_balanced_smote_2)] <- "Risk.Rating"
df_balanced_smote_2$Risk.Rating <- factor(df_balanced_smote_2$Risk.Rating,
                                          labels = levels(financial_df_cleaned$Risk.Rating))

# Step 5: Normalize features
df_balanced_smote_2[,-ncol(df_balanced_smote_2)] <- as.data.frame(
  lapply(df_balanced_smote_2[,-ncol(df_balanced_smote_2)], MinMax)
)

# Step 6: Train/Test split
set.seed(123)
train_idx <- sample(nrow(df_balanced_smote_2), 0.7 * nrow(df_balanced_smote_2))
train_data <- df_balanced_smote_2[train_idx, ]
test_data <- df_balanced_smote_2[-train_idx, ]

# Step 7: Cross-Validated Grid Search for Tuned KSVM
X_train <- train_data[, -ncol(train_data)]
y_train <- train_data$Risk.Rating
X_test <- test_data[, -ncol(test_data)]
y_test <- test_data$Risk.Rating

C_values <- c(0.1, 1, 10)
sigma_values <- c(0.01, 0.05, 0.1)

folds <- createFolds(y_train, k = 5)

best_acc <- 0
best_model <- NULL
best_C <- NA
best_sigma <- NA

for (C_val in C_values) {
  for (sigma_val in sigma_values) {
    accs <- c()
    
    for (fold in folds) {
      X_tr <- X_train[-fold, ]
      y_tr <- y_train[-fold]
      X_val <- X_train[fold, ]
      y_val <- y_train[fold]
      
      model <- ksvm(
        x = as.matrix(X_tr),
        y = y_tr,
        type = "C-svc",
        kernel = "rbfdot",
        C = C_val,
        kpar = list(sigma = sigma_val),
        scaled = TRUE
      )
      
      pred <- predict(model, as.matrix(X_val))
      accs <- c(accs, mean(pred == y_val))
    }
    
    mean_acc <- mean(accs)
    cat(sprintf("C=%.2f, sigma=%.3f --> CV Accuracy=%.4f\n", C_val, sigma_val, mean_acc))
    
    if (mean_acc > best_acc) {
      best_acc <- mean_acc
      best_model <- model
      best_C <- C_val
      best_sigma <- sigma_val
    }
  }
}

cat(sprintf("Best Model: C=%.2f, sigma=%.3f with CV Accuracy=%.4f\n", best_C, best_sigma, best_acc))

# Step 8: Final prediction on test set
svm_pred <- predict(best_model, as.matrix(X_test))
svm_pred <- factor(svm_pred, levels = levels(y_test))
truth <- y_test

# Step 9: Evaluation
confusionMatrix(svm_pred, truth)

# Per-class metrics
classes <- levels(truth)
for (cls in classes) {
  cat(paste0("\nClass: ", cls, "\n"))
  precision <- Precision(y_pred = svm_pred, y_true = truth, positive = cls)
  recall <- Recall(y_pred = svm_pred, y_true = truth, positive = cls)
  f1 <- F1_Score(y_pred = svm_pred, y_true = truth, positive = cls)
  
  cat(paste("  Precision:", round(precision, 4), "\n"))
  cat(paste("  Recall:   ", round(recall, 4), "\n"))
  cat(paste("  F1 Score: ", round(f1, 4), "\n"))
}

```


```{r}
# Retrain the best SVM model with probability estimation enabled
svm_model_prob <- ksvm(
  x = as.matrix(X_train),
  y = y_train,
  type = "C-svc",
  kernel = "rbfdot",
  C = best_C,
  kpar = list(sigma = best_sigma),
  prob.model = TRUE,
  scaled = TRUE
)

# Predict class probabilities on the test set
svm_prob <- predict(svm_model_prob, as.matrix(X_test), type = "probabilities")

# True labels and class levels
true_labels <- y_test
classes <- levels(true_labels)
colors <- c("red", "blue", "darkgreen")  # Assign a color to each class

# Plot empty ROC canvas
plot(NULL, xlim = c(0,1), ylim = c(0,1),
     xlab = "False Positive Rate", ylab = "True Positive Rate",
     main = "Tuned SVM ROC Curve")

# Plot ROC curve for each class (one-vs-rest)
for (i in seq_along(classes)) {
  cls <- classes[i]
  binary_labels <- as.numeric(true_labels == cls)
  
  pred <- prediction(predictions = svm_prob[, cls], labels = binary_labels)
  perf <- performance(pred, "tpr", "fpr")
  
  plot(perf, add = TRUE, col = colors[i], lwd = 2)
}

# Add diagonal reference line
abline(a = 0, b = 1, col = "gray", lty = 2)

# Add legend
legend("bottomright", legend = classes, col = colors, lty = 1, lwd = 2, bty = "n")

# Calculate and print AUC for each class
for (i in seq_along(classes)) {
  cls <- classes[i]
  binary_labels <- as.numeric(true_labels == cls)
  pred <- prediction(predictions = svm_prob[, cls], labels = binary_labels)
  auc <- performance(pred, "auc")@y.values[[1]]
  cat(paste0("AUC for class ", cls, ": ", round(auc, 4), "\n"))
}


```


## 3.2 XGboost

Assumption: Using df_balanced_smote_2 (after two rounds of SMOTE)

### Data Preparation
```{r}

# Ensure the target is a factor
df_balanced_smote_2$Risk.Rating <- as.factor(df_balanced_smote_2$Risk.Rating)

# Split the data (if not already done)
set.seed(1996)
train_idx <- sample(nrow(df_balanced_smote_2), 0.7 * nrow(df_balanced_smote_2))
train_data <- df_balanced_smote_2[train_idx, ]
test_data <- df_balanced_smote_2[-train_idx, ]


```
### Define Formula and Training Control
```{r}
xgb_formula <- Risk.Rating ~ .

ctrl <- trainControl(
  method = "cv", number = 5,
  classProbs = TRUE,
  summaryFunction = multiClassSummary,
  allowParallel = TRUE
)

```

### Train XGBoost Model

```{r}
set.seed(123)

xgb_model <- train(
  xgb_formula,
  data = train_data,
  method = "xgbTree",
  trControl = ctrl,
  tuneLength = 5,
  metric = "Accuracy"
)

```
### Model Evaluation
```{r}

# Predictions
xgb_pred <- predict(xgb_model, newdata = test_data)

# Confusion Matrix
confusionMatrix(xgb_pred, test_data$Risk.Rating)

# Precision / Recall / F1-Score


truth <- test_data$Risk.Rating
for (cls in levels(truth)) {
  cat(paste0("\nClass: ", cls, "\n"))
  precision <- Precision(y_pred = xgb_pred, y_true = truth, positive = cls)
  recall <- Recall(y_pred = xgb_pred, y_true = truth, positive = cls)
  f1 <- F1_Score(y_pred = xgb_pred, y_true = truth, positive = cls)
  
  cat(paste("  Precision:", round(precision, 4), "\n"))
  cat(paste("  Recall:   ", round(recall, 4), "\n"))
  cat(paste("  F1 Score: ", round(f1, 4), "\n"))
}


```

Show ROC Curve


```{r}

# Predict class probabilities
xgb_prob <- predict(xgb_model, newdata = test_data, type = "prob")
head(xgb_prob)  # Should contain 3 columns: High, Low, Medium

true_labels <- test_data$Risk.Rating
classes <- levels(true_labels)
colors <- c("red", "blue", "darkgreen")

# Create empty plot
plot(NULL, xlim = c(0, 1), ylim = c(0, 1), xlab = "False Positive Rate", ylab = "True Positive Rate",
     main = "XGBoost ROC Curve - One vs Rest")

# Loop through each class
for (i in seq_along(classes)) {
  cls <- classes[i]
  
  # One-vs-Rest binary labels
  binary_labels <- as.numeric(true_labels == cls)
  
  # Get probabilities for class i
  pred <- prediction(predictions = xgb_prob[, cls], labels = binary_labels)
  perf <- performance(pred, "tpr", "fpr")
  
  # Plot curve
  plot(perf, add = TRUE, col = colors[i], lwd = 2)
}

abline(a = 0, b = 1, col = "gray", lty = 2)
legend("bottomright", legend = classes, col = colors, lty = 1, lwd = 2, bty = "n")


```

Calculate AUC
```{r}
for (i in seq_along(classes)) {
  cls <- classes[i]
  binary_labels <- as.numeric(true_labels == cls)
  pred <- prediction(xgb_prob[, cls], binary_labels)
  auc <- performance(pred, "auc")@y.values[[1]]
  cat(paste0("AUC for class ", cls, ": ", round(auc, 4), "\n"))
}
```
